\chapter{SAM/BAM files}

\section{The SAM/BAM file format}
\begin{information}
Reads that have been to a reference are no longer stored in fastq format but are stored in either SAM or BAM format.
These two formats are virtually identical, however the SAM format is a text file which is easily readable to human eyes, whilst a BAM file is the same information converted to binary.
This conversion means that file sizes are smaller, and that computational processes can be performed more efficiently.
Typically, we work with BAM files as these provide gains in storage space \& analytic speed.
The tools we use to inspect these files are provided in the package samtools, which has been installed on your VM. \\
\end{information}

\begin{steps}
The reads from the previous dataset which mapped to \textit{chr2L} of \textit{D. melanogaster} are in the folder \texttt{\~{}/mapped}.
Let's look at the first 10 lines of the file.
\begin{lstlisting}
cd ~/mapped
head chr2L.sam  
\end{lstlisting}
\end{steps}

\begin{questions}
Can you distinguish between the header of the SAM format and the actual alignments?
\begin{answer}
The header line starts with the letter `@', i.e.: \\
\begin{tabular}{llll}
@HD & VN:1.0 & SO:unsorted & \\
@SQ & SN:chr2L & LN:23011544 & \\
@PG & ID:TopHat	& VN:1.2.0 & CL:/g/steinmetz/collaboration/... \\
\end{tabular}

While the actual alignments start with read id, i.e.: \\

\begin{tabular}{llll}
SRR031714.5049824 & 99 & chr2L & etc \\
SRR031714.5049824 & 147 & chr2L & etc \\
\end{tabular}

\end{answer}
What kind of information does the header provide? \\
\begin{answer}
\begin{itemize}
\item @HD: Header line; VN: Format version; SO: the sort order of alignments.
\item @SQ: Reference sequence information; SN: reference sequence name; LN: reference sequence length.
\item @PG: Read group information; ID: Read group identifier; VN: Program version; CL: the command line that produces the alignment.
\end{itemize}
\end{answer}
\end{questions}

\section{Conversion to BAM format}
\begin{steps}
The BAM format is much more convenient computationally, so we can convert this file into the BAM format using \texttt{samtools view}.
Whilst this line of code may look counter-intuitive, remember that the first section is all the information about how to process the input, \& specify the output, whilst the final section is the file that the command will operate on.
First look through the help page to understand what you are asking the tool to perform. \\
\begin{lstlisting}
samtools view 
samtools view -bSo chr2L.bam chr2L.sam
\end{lstlisting}
Have a look at the file sizes to see the difference.
Note that the BAM file is about 12\% of the size, which can be an important consideration when managing large datasets.
\begin{lstlisting}
ls -lh
\end{lstlisting}
\end{steps}

\section{Handling BAM files}
\begin{steps}
As BAM files are in binary format they will look like gibberish if we try to read them directly.
Instead we can inspect them by using \texttt{samtools view} again.
To view the header we use:\\
\begin{lstlisting}
samtools view -H chr2L.bam
\end{lstlisting}
Whereas to look through the first few mapped reads, we can use:
\begin{lstlisting}
samtools view chr2L.bam | head
\end{lstlisting}
As this data can easily spill across lines, it might be helpful to maximise your terminal to see the complete line structure.
\end{steps}

\subsection{The SAM/BAM data structure}
\begin{information}
If we understand what information is contained within a file, we can know what decisions to make as we progress with our analysis, so let's have a look at what the data structure is for a SAM/BAM file.
A SAM file is \textit{tab-delimited}, which means that each field is separated by a tab, giving a data structure effectively consisting of columns (or fields).
In order, these are: \\
\begin{tabular}{r l l}
1 & QNAME & Query template/pair NAME \\
2 & FLAG & bitwise FLAG \\
3 & RNAME & Reference sequence NAME \\
4 & POS & 1-based leftmost POSition/coordinate of clipped sequence \\
5 & MAPQ & MAPping Quality (Phred-scaled) \\
6 & CIGAR & extended CIGAR string \\
7 & MRNM & Mate Reference sequence NaMe (`=' if same as RNAME) \\
8 & MPOS & 1-based Mate POSistion \\
9 & TLEN & inferred Template LENgth (insert size) \\
10 & SEQ & query SEQuence on the same strand as the reference \\
11 & QUAL & query QUALity (ASCII-33 gives the Phred base quality) \\
12+ & OPT & variable OPTional fields in the format TAG:VTYPE:VALUE \\
\end{tabular}
\end{information}

\begin{note}
Several of these fields contain useful information, so looking the the first few lines which we displayed above, you can see that these reads are mapped in pairs as consecutive entries in the QNAME field are often (but not always) identical.
Most of these fields are self-explanatory, but some require exploration in more detail.
\end{note}

\subsection{SAM Flags}
\begin{information}
These are quite useful pieces of information, but can be difficult at first look.
Head to \url{http://broadinstitute.github.io/picard/explain-flags.html} to see a helpful description.
The simplest way to understand these is that it is a bitwise system so that each description heading down the page increases ina binary fashion.
The first has value 1, the second has value 2, the third has value 4 \& so on until you reach the final value of 2048.
The integer value contained in this file is the unique sum of whichever attributes the mapping has.
For example, if the read is paired \& mapped in a proper pair, but no other attributes are set, the flag field would contain the value 3.
\end{information}

\begin{questions}
What value could a flag take if the read was 1 - paired; 2 - mapped in a proper pair; 3 - it was the first in the pair \& 4 - the alignment was a supplementary alignment. \\
\begin{answer}
2115 \\
\end{answer}
Some common values in the bam file are 99, 147 \& 145.
Look up the meanings of these values. \\
\end{questions}

\begin{information}
Things can easily begin to confuse people once you start searching for specific flags, but if you remember that each attribute is like an individual flag that is either on or off (i.e. it is binary).
If you searched for flags with the value 1, you wouldn't obtain the alignments with the exact value 1, rather you would obtain the alignments for which the first flag is set \& these can take a range of values.
\end{information}

\begin{steps}
Let's try this using the command \texttt{samtools view} with the option \texttt{-f N} to include reads with a flag set and the option \texttt{F N} to exclude reads with a specific flag set.
Let's get the first few reads which are mapped in a proper pair, so the flag `2' will be set.\\
\begin{lstlisting}
samtools view -f 2 chr2L.bam | head
\end{lstlisting}
Note that none of the flags actually have the value 2, but if you typed the values 99, 147 or 163 into the webpage, you'll see that this flag is set for all of these values.
Similarly if we wanted to extract only the reads which are NOT mapped in a proper pair we would change the option to a upper-case F.
\begin{lstlisting}
samtools view -F 2 chr2L.bam | head
\end{lstlisting}
Again, try entering a few of these sample values into the webpage and you will see that this flag is not set for any of these values.
\end{steps}

\begin{information}
This can be a very helpful tool for extract subsets of your aligned reads.
For example, we can create a new BAM file with only the reads which were aligned in a proper pair by entering the following command.
\begin{lstlisting}
samtools view -f 2 -bo chr2L.paired.bam chr2L.bam
ls -lh
\end{lstlisting}
You can pull out highly specific combinations of alignments should you so choose
\end{information}

\subsection{CIGAR strings}
\begin{note}
These give useful information about the type of alignment that has been performed on the read.
In the first few reads we called up earlier, most had the value \texttt{37M}, which stands for 37 \textbf{M}atches.
The other abbreviations in common use are I (insertion) \& D (deletion), however these are not found in this bam file.
\end{note}

\begin{questions}
A CIGAR string contained in this file is `10M240N27M'.
Can you think of an interpretation of this string? \\
\begin{answer}
10 Matches, 240 Skipped bases in the reference, then 27 Matches.
This may represent a splice junction as determined by \texttt{tophat}
\end{answer}
\end{questions}

\begin{advanced}
Knowing this pattern as splice junctions, you could write a \textit{regular expression} search to find only the alignments that contain potential splice junctions.
\begin{questions}
What would be a pattern to search for, and how would you create this as a new BAM file.
\end{questions}
\begin{answer}
This will pull out the first few.
\begin{lstlisting}
samtools view chr2L.bam | egrep '[0-9]+M[0-9]+N[0-9]+M' | head
\end{lstlisting}
To really do it correctly, you'd probably have to fudge a new sam file with the header, the cat the output \& convert to BAM
\end{answer}
\end{advanced}